---
title: "COVID Wastewater Surveillance for Western Massachusetts"
output:
  html_document:
    df_print: paged
---

```{r Setup, echo=FALSE, message=FALSE, warning=FALSE}

```

```{r Download and process, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(here)
library(lubridate)
library(ggthemes)
library(scales)
library(patchwork)
library(readxl)
library(ggtext)
options(scipen = 999)
source("wva_calculator.R")
source("mytheme.R")

### Download and initial cleaning ----

url_current <- "https://www.mass.gov/doc/wastewater-dashboard-data/download"
url_arch <- "https://www.mass.gov/doc/2023-2024-archived-wastewater-dashboard-data/download"

download.file(url_current, destfile = "sample_current.xlsx", mode = "wb")
download.file(url_arch, destfile = "sample_arch.xlsx", mode = "wb")

raw_sample_current <-
  read_excel(path = "sample_current.xlsx", sheet = "Wastewater Testing Data") %>%
  rename_all(~ str_to_lower(.)) %>%
  rename_all( ~ str_replace_all(., "[ ]", "_")) %>% 
  mutate(date = ymd(sample_collection_date), source="current") %>%
  write_csv(paste0(here("data_downloads", paste0(Sys.Date(), "_sample_current.csv"))))

raw_sample_arch <-
  read_excel(path = "sample_arch.xlsx", sheet = "Wastewater Testing Data") %>%
  rename_all(~ str_to_lower(.)) %>%
  rename_all( ~ str_replace_all(., "[ ]", "_")) %>% 
  mutate(date = ymd(sample_collection_date), source="arch") %>%  
  write_csv(paste0(here("data_downloads", paste0(Sys.Date(), "_sample_arch.csv"))))

weights <- read.csv("better_pop_weights_Aug2024.csv") %>%
  select(-mgd) %>% rename(pop_weight = users) %>% 
  mutate(pop_weight = as.numeric(pop_weight))

wmass_counties <- c("Hampden", "Franklin", "Hampshire", "Berkshire")
nvalley_counties <- c("Franklin", "Hampshire")
svalley_counties <- c("Hampden", "Hampshire")
berkshire <- "Berkshire"

### Set time parameter
time_years = 1

### Prepare the data
full_sample <-
  # combine archive and current data
  bind_rows(raw_sample_arch %>% 
              filter(date < min(raw_sample_current$date)),
            raw_sample_current %>% select(-respiratory_season)) %>% 
  # rename columns 
  rename(
    "avg_7d_covid_conc" = '7_day_average_of_sars-cov-2_concentration',
    "county" = county_of_sampling_location,
    "city_sample" = city_of_sampling_location,
    "system" = name_of_sampling_location) %>%
  # convert to copies / mL
  mutate(avg_7d_covid_conc = avg_7d_covid_conc/1000) %>% 
  # modify for log-scale axis display by adding small constant to zero values
  mutate(avg_7d_covid_conc = ifelse(avg_7d_covid_conc == 0, .01, avg_7d_covid_conc)) %>%
  # remove unused rows
  select(-c(tester, site_type, 'units_(measuring_concentration)', state_of_sampling_location)) %>%
  # reorganize columns
  relocate(date) %>% relocate(city_sample, .after = date) %>% relocate(county, .after = date) %>%
  # shift to uniform lower case
  mutate(city_sample = str_to_lower(city_sample),
         # create lowercase version of system for easy join to pop weights
         weights_id = str_to_lower(system)) %>%
  # add geographical groupings
  mutate(
    wmass = county %in% wmass_counties,
    nvalley = county %in% nvalley_counties,
    svalley = county %in% svalley_counties) %>%
  # remove extraneous columns
  select(date, county, system, avg_7d_covid_conc, weights_id, wmass, nvalley, svalley) %>%
  # add population weights
  left_join(weights, by = join_by(weights_id == system)) %>%
  select(-weights_id) %>%
  # calculate population-days as weights for loess fit
  # first calculate reporting interval i.e. days til next report for that location
  arrange(date, .by=system) %>%
  mutate(
    interval = ((lead(date, default = max(date)+ddays(1)))-date)/ddays(1), .by=system,
    # interval multiplier floor is at 0.5 (for the max of 2 entries in a single day
    # ceiling is at 7, so as to not over-weight systems that miss time
    pop_day = pop_weight * (min(interval, 7) %>% max(0.5))) %>%
  ## label runs between gaps for each system
  group_by(system) %>%
  mutate(
    gap = if_else(date > dplyr::lag(date, default=lubridate::origin)+dweeks(2), 1, 0),
    run=cumsum(gap),
    gap=NULL,
    smooth_group = paste0(system, "_", run)) %>% ungroup() %>%
  arrange(system, date) 

# save(full_sample, weights, file="current.Rda")

wva_pal <- c("#49a667", "#b6c459", "#8fa1ef", "#fdae61", "#f45489")
time_years = 1

### -----

### Load data ----
# load("current.Rda")

```


```{r Prepare regional plot, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Generate smooth and wvas
# smooth_operator automaticaly filters for western mass
# we can customize that for future iterations
dfreg <- smooth_operator(full_sample, time_years=1)
thresholds_reg <- wtd_ww_threshold_breaks(dfreg$avg_7d_covid_conc, wts=dfreg$pop_day)

```


```{r Construct regional plot, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
### construct regional plot
region <- 
  dfreg %>%
  ggplot() +
  geom_hline(yintercept = thresholds_reg[2:5], color="grey50", linetype="dotted") +
  geom_point(aes(date, avg_7d_covid_conc, size=pop_day, color=wva_point), alpha=.35) +
  geom_line(aes(date,  y_smoothed, group=1), color="darkblue", linewidth=1.25) +
  scale_y_log10(limits = c(10, 8000)) + 
  scale_color_manual(values = wva_pal) +
    labs(x = "Date",
         y = "7-day COVID Concentration (copies/mL)",
         color= "Wastewater Viral\nActivity Level") +
    guides(color = guide_legend(reverse=TRUE, override.aes = list(alpha = 1)),
           fill = guide_legend(reverse=TRUE),
           size="none") +
  mytheme()

plot_title_reg <- ifelse(time_years == 1, 
                     paste0("wma_wastewater_", max(dfreg$date), "_region",  ".tiff"), 
                     paste0("wma_wastewater_", max(dfreg$date), "_region_", time_years, "YEAR.tiff"))

width <- ifelse(time_years == 1, 2100, 3300)
height <- 1900

ggsave(region, path = here("plots/"), filename = plot_title_reg,
       width = width, height = height, units = "px", bg="white")
```
 
 
 <font size="5">Regional Concentration and Wastewater Viral Activity Level</font>
```{r Display summary message, echo=FALSE, message=FALSE, warning=FALSE, comment=NA}
subtitle <- paste0("From ", min(dfreg$date), " to ", max(dfreg$date),
                   "\nCurrent WVA level: ",  dfreg %>% arrange(date) %>% pull(wva_lvl) %>%  last() %>% as.character(),
                   "\nCurrent WVA score: ",  dfreg %>% arrange(date) %>% pull(wva_score) %>%  last() %>% as.character())
cat(subtitle)
```
```{r Define sewer number, echo=FALSE, warning=FALSE, comment=NA}
sewer_num <- length(dfreg %>% pull(system) %>% unique())
```


```{r Plot region, echo=FALSE, message=F, fig.height=4, fig.width=8, message=FALSE, warning=FALSE, fig.cap=sprintf("The above figure shows the regional wastewater COVID concentration for %d sewer systems in Berkshire, Hampden, Hampshire, and Franklin counties. The trend line is a smoothed conditional mean and each point is an individual system report. Trend line and individual reports are each weighted by population and reporting interval. WVA levels were calculated using the CDC methodology, except the (%d year) baseline is updated each time the WVA is calculated.", sewer_num, time_years)}
plot(region)
```


 <font size="5">County-level Concentration and Wastewater Viral Activity Level</font>
```{r Prepare county plot, echo=FALSE, warning=FALSE}
counties <- full_sample %>% filter(date > max(date) - years(time_years) & wmass) %>% pull(county) %>% unique()

# create the data.frame to hold the county smooths
dfcount <- data.frame(matrix(ncol = ncol(full_sample)))[-1,]
colnames(dfcount) <- names(full_sample)
summ <- data.frame(county = NA, note = NA)[-1,]
# loop through counties
for (i in seq_along(counties)){
  # create smooths
  temp <- smooth_operator(full_sample %>% filter(county == counties[i]), time_years=1)
  dfcount <- rbind(dfcount, temp)
  # create summary annotations
  tempnote <- data.frame(county = counties[i], 
                         note = paste0(max(temp$date),"\nWVA level: ",  temp %>% arrange(date) %>% pull(wva_lvl) %>%  last() %>% as.character(),"\nWVA score: ",  temp %>% arrange(date) %>% pull(wva_score) %>%  last() %>% as.character()))
  summ <- rbind(summ, tempnote)
  # create dataframe of thresholds to pass to hline
}
# add xy coordinates to annotations
summ$x <- min(temp$date)
summ$y <- 11

#### county plot ----
county_plot <-
dfcount %>% 
  ggplot() +
  geom_point(aes(date, avg_7d_covid_conc, size=pop_day, color=wva_point), alpha=.35) +
  geom_line(aes(date,  y_smoothed, group=1), color="darkblue", linewidth=1.15) +
  scale_y_log10(limits = c(10, 8000)) + 
  scale_color_manual(values = wva_pal) +
  guides(color = guide_legend(reverse=TRUE, override.aes = list(alpha = 1)),
         fill = guide_legend(reverse=TRUE),
         size="none") +
  labs(x = "Date",
       y = "7-day COVID Concentration (copies/mL)",
       color= "Wastewater Viral\nActivity Level") +
  geom_label(data = summ,
             aes(x=x, y=y, label = note), color="blue", hjust=0,vjust=0,
             size=2.5) + 
  mytheme() + # set the left align here +
  theme(plot.title = element_text(size=12)) +
  facet_wrap(~county, ncol=2)

plot_title_count <- paste0("wma_wastewater_", max(dfreg$date), "_counties", ".tiff")
width = 1992*1.25
height = 1200*1.25

ggsave(county_plot, path = here("plots/"), filename = plot_title_count,
       width = width, height = height, units = "px", bg="white")

```
```{r Plot counties, echo=FALSE, fig.height=6.25, fig.width=10, message=FALSE, warning=FALSE, fig.cap = paste0("County-level wastewater COVID concentration for ", sewer_num, " sewer systems in Berkshire, Hampden, Hampshire, and Franklin counties. Trend line is smoothed conditional mean, points are individual system reports. Both are weighted by population and reporting interval.")}
plot(county_plot)
```


```{r Prepare Systems Plot, echo=FALSE, message=FALSE, warning=FALSE}
# title_sys <- "Wastewater COVID Concentration for Sewer Systems in Western MA"
# subtitle_sys <- paste0(min(dfreg$date), " to ", max(dfreg$date))
# caption_sys <- "Trend line is smoothed conditional mean, points are individual system reports. Both are weighted by population and reporting interval.\nBe careful out there <3 western.ma.covid.data@gmail.com"

smooth_groups <- full_sample %>% 
  filter(date > max(date) - months(6) & wmass) %>% select(system, group=smooth_group) %>% unique()

# create the data.frame to hold the system smooths
dfsys <- data.frame(matrix(ncol = ncol(full_sample)))[-1,]
colnames(dfsys) <- names(full_sample)
summ_temp <- data.frame(max_date = NA, system = NA, smooth_group = NA, note = NA)[-1,]
# loop through counties
for (i in seq_along(smooth_groups$group)){
  # create smooths
  temp <- smooth_operator(full_sample %>% filter(date > max(date) - years(time_years)) %>% filter(smooth_group == smooth_groups$group[i]), time_years=1)
  # qc <- temp %>% select(date, smooth_group)
  dfsys <- rbind(dfsys, temp)
  # create summary annotations by system
  tempnote <- data.frame(max_date = max(temp$date),
                         system = smooth_groups$system[i], 
                         smooth_group = smooth_groups$group[i],
                         note = 
                           paste0(max(temp$date),"\nWVA level: ",  
                                  temp %>% arrange(date) %>% pull(wva_lvl) %>%  last() %>% as.character(),"\nWVA score: ",  temp %>% arrange(date) %>% pull(wva_score) %>%  last() %>% as.character()))
  summ_temp <- rbind(summ_temp, tempnote)
  # create dataframe of thresholds to pass to hline
}
# extract most recent system level information (discarding update from older runs)
summ <- summ_temp %>% group_by(system) %>% slice_max(max_date, n=1)

#add xy coordinates to annotations
summ$x <- min(full_sample %>% 
                filter(date > max(date) - years(time_years) & wmass) %>% pull(date))
summ$y <- 11

#### systems plot ----
systems_plot <- 
  dfsys %>% 
  ggplot() +
  geom_line(aes(date,  y_smoothed, group=smooth_group), color="darkblue", linewidth=1.1) +
  geom_point(aes(date, avg_7d_covid_conc, color=wva_point), alpha=.35) +
  scale_y_log10(limits = c(10, 8000)) + 
  scale_color_manual(values = wva_pal) +
  guides(color = guide_legend(reverse=TRUE, override.aes = list(alpha = 1)),
         fill = guide_legend(reverse=TRUE),
         size="none") +
  labs(x = "Date",
       y = "7-day COVID Concentration (copies/mL)",
       color= "Wastewater Viral\nActivity Level") +
  geom_label(data = summ,
             aes(x=x, y=y, label = note), color="blue", hjust=0,vjust=0,
             size=2.5) + 
  mytheme() + 
  theme(plot.title = element_text(size=12)) +
  facet_wrap(~system, ncol=3) +
    scale_x_date(date_labels = "%b\n%Y")
```
 <font size="5">Concentration for Individual Sewer Systems and Wastewater Viral Activity Level</font>
```{r Systems plot, echo=FALSE, message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.cap = paste0("The above figure shows reported concentration data from individual systems, with linear interpolation between reports of varying frequency. Gaps of longer than one week are not interpolated. Blue annotations indicate date of last data reported.")}
plot(systems_plot)
```

 <font size="3">**Data Sources**</font>  
• CDC methodology: https://www.cdc.gov/nwss/about-data.html#data-method  
• Wastewater data access: https://www.mass.gov/info-details/wastewater-surveillance-reporting  
• Population weights are from various sources (EPA estimates whenever available).  
<br>  
<br>  
Be careful out there <3 [western.ma.covid.data@gmail.com]
